---
title: "2 Discrete random variable"
author: "SUNC"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: default
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: no
  word_document:
    toc: yes
    toc_depth: '3'
  pdf_document:
    toc: yes
    toc_depth: '3'
    latex_engine: xelatex
header-includes:
- \usepackage{xeCJK}
- \usepackage{fontspec}
- \setCJKmainfont{微軟正黑體}
- \XeTeXlinebreaklocale "zh"
- \XeTeXlinebreakskip = 0pt plus 1pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(corrplot)
library(ggplot2)
library(tidyverse)
```


[chatgpt](https://chatgpt.com/share/67f8cba3-0e98-8003-ad1b-44369d30f924)

A random variable \( Y \) is a function that maps outcomes from the sample space  
\( S \) of a random experiment to real numbers.

- If \( Y \) only takes integer values, it is called a **discrete random variable**.

- The **probability mass function (PMF)** is denoted by \( P(Y = x),\ x \in \mathbb{N} \).

- The **cumulative distribution function (CDF)** is denoted by \( P(Y \leq x),\ x \in \mathbb{N} \).

- The **expectation** and **variance** of \( Y \) are denoted by \( E(Y) \) and \( \text{Var}(Y) \), respectively.

  

## **Statistic vs. Estimate**

A **random sample** of size \( n \) from \( Y \) is denoted as:

\[
Y_1, Y_2, \dots, Y_n.
\]

- A **statistic** is any function of the sample, such as \( \bar{Y}_n \)
  which estimates the population mean \( E(Y) \).

- An **estimate** is a statistic used to infer the value of an unknown parameter.


## **Example**

- Example 2.1-3, *Hogg, Tanis and Zimmerman, 9th Edition*, p.44:  
  Roll a fair four-sided die twice and let \( X \) be the maximum of the two outcomes.


## **1 基於公正骰子的假設**

  - \( P(X = k) = \frac{2k - 1}{16}, \quad k = 1, 2, 3, 4. \)
  
  - 期望值 (Expectation): \( E(X) = \frac{25}{8} \approx 3.125. \)
    
  - 變異數 (Variance): \( \text{Var}(X) = \frac{55}{64} \approx 0.8594,\ \text{SD}(X) \approx 0.927. \)
  
---

## **2 樣本觀測與統計量**

從此分佈抽取 \( n \) 個樣本 \( x_1, x_2, \dots, x_n \)，可以計算以下 **7 個統計量**：

- **機率估計**： $\hat{P}(X = k) = \frac{1}{n} \sum_{i=1}^{n} I_{\{x_i = k\}}, \quad k = 1,2,3,4.$

- **樣本平均 (Sample Mean)**：$\bar{x} = \frac{1}{n} \sum_{i=1}^{n} x_i.$

- **樣本變異數 (Sample Variance)**：$s^2 = \frac{1}{n-1} \sum_{i=1}^{n} (x_i - \bar{x})^2.$

- **卡方統計量 (Chi-Square Statistic)**：(綜合衡量機率估計與預期機率的差距)
  \[
  \chi^2 = \sum_{k=1}^{4} \frac{(o_k - E_k)^2}{E_k},
  \]
  
  - \( o_k \) 是第 \( k \) 種結果的觀測次數 $n \hat{P}(X = k)$,
  
  - \( E_k \) 是第 \( k \) 種結果的期望次數 \( n P(X = k) \).


```{r, echo=FALSE}
par(mfrow=c(1,2))

n <- 20
dicewt <- c(1,1,1,1)  # 骰子權重
dicep <- prop.table(dicewt)   # 骰子機率
p0.ss <- rep(1/16,length(dicewt)^2)  # Sample Space H0:公正骰子所投的結果
p0.max <- (2*(1:4)-1)/16 # Max H0:公正骰子所投的結果

ds <- as.data.frame(matrix(sample(1:4, 2*n, replace=TRUE, prob=dicep), ncol=2))
colnames(ds) <- c("Var1", "Var2")  # 設定欄位名稱

# 使用 transform() 將變數轉為 factor，確保所有可能組合都出現，並加入 Max 欄位
ds <- transform(ds, Var1 = factor(Var1, levels = 1:4), 
                     Var2 = factor(Var2, levels = 1:4), 
                     Max  = pmax(as.numeric(Var1), as.numeric(Var2)))

fq.ss <- as.vector(table(ds$Var1, ds$Var2))  
plot(prop.table(fq.ss), type = "h", ylab = "", xlab = "Sample Space",
     ylim = c(0, 0.1), xaxt = "n", col = "#0000FF")
abline(h=p0.ss[1],col="#FF0000")
chistat.ss <- round(sum((fq.ss-n*p0.ss)^2/(n*p0.ss)),4)
legend("topright",c("Avg = ?","Sdv = ?",paste("Diff =",chistat.ss)))
# chisq.test(fq.ss,p=p0.ss) # 卡方適合度檢定

# r.v. X的各種各能的相對頻率
fq.max <- table(factor(ds$Max,levels=1:4))
plot(p0.max,type="h",lwd=2,xlim=c(0,length(dicewt))+0.5,ylim=c(0,1),ylab="",col="#FF0000",xlab="Max")
points(prop.table(fq.max),type="h",lwd=10,col="#0000FF50")
chistat.max <- sum((fq.max-n*p0.max)^2/(n*p0.max))
legend("topright",
       paste(c("Avg =","Sdv =","Diff =" ),
             round(c(mean(ds$Max),sd(ds$Max),chistat.max),4)))
# chisq.test(table(ds$Max),p=p0.max) # 卡方適合度檢定

mtext(paste("Sample Size: n =", n), outer=TRUE, line=-2)
mtext(paste(c("Dice Prob = (",paste(dicep,collapse=", "),")"),collapse=""), outer=TRUE, line=-1)
```

- 先撰寫一個 視覺化版本（非互動式），觀察樣本數變化時，
模擬統計量（如平均、標準差）與理論值之間的 誤差趨勢。

- 將上述內容擴充為 Shiny App 的互動版本，讓使用者可以自由調整**樣本數**與**骰子權重**，
並即時觀察絕對差的變化。


---

## **3 統計量也是隨機變數**

- \( Y_1, \cdots, Y_n \) is a random sample of \( Y \). Hence, \( h(Y_1, \cdots, Y_n) \) is a random variable,  
  where \( h \) is a function that maps the sample to a real number.

  - \( \hat{P}(X = k) = \frac{1}{n} \sum_{i=1}^{n} I_{\{X_i = k\}}, \quad k = 1, 2, 3, 4. \)

  - \( \bar{X} = \frac{1}{n} \sum_{i=1}^{n} X_i. \)

  - \( S^2 = \frac{1}{n-1} \sum_{i=1}^{n} (X_i - \bar{X})^2. \)


```{r}
p0.max <- (2*(1:4)-1)/16 # Max H0:公正骰子所投的結果

simulate_max_summary <- function(n = 100, dicewt = c(1, 1, 1, 1), seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  dicep <- prop.table(dicewt)

  ds <- as.data.frame(matrix(sample(1:4, 2 * n, replace = TRUE, prob = dicep), ncol = 2))
  colnames(ds) <- c("Var1", "Var2")
  ds$Max <- pmax(ds$Var1, ds$Var2)

  ds$max_mean <- NA
  ds$max_var  <- NA
  ds$max_sd   <- NA
  ds$max_1 <- NA
  ds$max_2 <- NA
  ds$max_3 <- NA
  ds$max_4 <- NA

  for (i in 1:n) {
    current <- ds$Max[1:i]
    ds$max_mean[i] <- mean(current)
    ds$max_var[i]  <- if (i > 1) var(current) else NA
    ds$max_sd[i]   <- if (i > 1) sd(current)  else NA

    fq <- table(factor(current, levels = 1:4))
    ds$max_1[i] <- fq[1] / i
    ds$max_2[i] <- fq[2] / i
    ds$max_3[i] <- fq[3] / i
    ds$max_4[i] <- fq[4] / i
  }

  return(ds)
}

# Example usage and plotting
result <- simulate_max_summary(n = 500, dicewt = c(1, 1, 1, 1), seed = 123)

par(mfrow = c(2, 1), mai=c(0.6, 0.8, 0.1, 0.2))

matplot(1:nrow(result), result[, c("max_1", "max_2", "max_3", "max_4")],
        type = "l", lty = 1, lwd = 2, xlim=c(2,nrow(result)), ylim = c(0,0.8), 
        col = adjustcolor(c("red", "blue", "darkgreen", "orange"), alpha.f = 0.6),
        ylab = "Relative Frequency", xlab = "Sample Size")

# 加上理論機率的虛線
p_theory <- p0.max
for (i in 1:4) {
  abline(h = p_theory[i], col = adjustcolor(c("red", "blue", "darkgreen", "orange")[i], alpha.f = 0.8), lty = 2)
}

legend("topright", legend = paste("Max =", 4:1),
       col = rev(adjustcolor(c("red", "blue", "darkgreen", "orange"), alpha.f = 0.6)), lty = 1)


plot(result$max_mean, type = "l", col = "blue", ylim = c(2.5, 3.5), xlim=c(2,nrow(result)),
     ylab = "Sample Mean", xlab = "")
abline(h = 25 / 8, col = "red", lty = 2)

plot(result$max_var, type = "l", col = "darkgreen", ylim = c(0.3, 1.2),xlim=c(2,nrow(result)),
     ylab = "Sample Variance", xlab = "")
abline(h = 55 / 64, col = "red", lty = 2)

plot(result$max_sd, type = "l", col = "purple", ylim = c(0.3, 1.2),xlim=c(2,nrow(result)),
     ylab = "Sample SD", xlab = "Sample Size")
abline(h = sqrt(55 / 64), col = "red", lty = 2)

```

```{r}
simulate_max_summary <- function(n = 100, dicewt = c(1, 1, 1, 1), r = 200, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  dicep <- prop.table(dicewt)

  # 初始化儲存 r 次模擬的結果列表
  result_list <- vector("list", r)

  for (rep in 1:r) {
    ds <- as.data.frame(matrix(sample(1:4, 2 * n, replace = TRUE, prob = dicep), ncol = 2))
    colnames(ds) <- c("Var1", "Var2")
    ds$Max <- pmax(ds$Var1, ds$Var2)

    # 初始化欄位
    ds$max_mean <- NA
    ds$max_var  <- NA
    ds$max_sd   <- NA
    ds$max_1 <- NA
    ds$max_2 <- NA
    ds$max_3 <- NA
    ds$max_4 <- NA

    for (i in 1:n) {
      current <- ds$Max[1:i]
      ds$max_mean[i] <- mean(current)
      ds$max_var[i]  <- if (i > 1) var(current) else NA
      ds$max_sd[i]   <- if (i > 1) sd(current)  else NA

      fq <- table(factor(current, levels = 1:4))
      ds$max_1[i] <- fq[1] / i
      ds$max_2[i] <- fq[2] / i
      ds$max_3[i] <- fq[3] / i
      ds$max_4[i] <- fq[4] / i
    }

    result_list[[rep]] <- ds
  }

  return(result_list)
}

```

- Usually, it is not easy to derive the distribution of \( h(Y_1, \cdots, Y_n) \), even if we know the distribution of \( Y \).

- If it is not feasible to derive the distribution analytically, we can approximate it by repeating the sampling process \( n_rep \) times, namely generating \( n_rep \) values of \( h(Y_1, \cdots, Y_n) \).

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(tibble)
  library(forcats)
  library(stringr)
  library(readr)
  library(lubridate)
})


# 理論機率（公正骰子取兩次最大值）
p0.max <- (2 * (1:4) - 1) / 16
mean_theory <- 25 / 8
var_theory <- 55 / 64
sd_theory <- sqrt(var_theory)

# 模擬函數（略作修改，回傳tibble）
simulate_max_summary <- function(n = 100, dicewt = c(1, 1, 1, 1)) {
  dicep <- prop.table(dicewt)
  ds <- tibble(
    Var1 = sample(1:4, n, replace = TRUE, prob = dicep),
    Var2 = sample(1:4, n, replace = TRUE, prob = dicep)
  )
  ds <- ds %>% mutate(Max = pmax(Var1, Var2))

  result <- tibble(
    SampleSize = 1:n,
    max_mean = NA_real_,
    max_var  = NA_real_,
    max_sd   = NA_real_,
    max_1 = NA_real_,
    max_2 = NA_real_,
    max_3 = NA_real_,
    max_4 = NA_real_
  )

  for (i in 1:n) {
    current <- ds$Max[1:i]
    fq <- table(factor(current, levels = 1:4))
    result$max_mean[i] <- mean(current)
    result$max_var[i]  <- if (i > 1) var(current) else NA_real_
    result$max_sd[i]   <- if (i > 1) sd(current)  else NA_real_
    result$max_1[i] <- fq[1] / i
    result$max_2[i] <- fq[2] / i
    result$max_3[i] <- fq[3] / i
    result$max_4[i] <- fq[4] / i
  }

  return(result)
}


n_rep <- 80
n <- 100
sim_all <- map_dfr(1:n_rep, ~simulate_max_summary(n), .id = "Rep") %>%
  filter(SampleSize > 10)


# 將資料整形為長格式
sim_long <- sim_all %>%
  pivot_longer(cols = starts_with("max_"), names_to = "Metric", values_to = "Value") %>%
  mutate(Metric = factor(Metric, levels = c("max_1", "max_2", "max_3", "max_4", "max_mean", "max_var", "max_sd")))

# 繪圖：相對頻率 (max_1~max_4)
sim_long %>%
  filter(Metric %in% c("max_1", "max_2", "max_3", "max_4")) %>%
  ggplot(aes(x = SampleSize, y = Value, group = Rep)) +
  geom_line(alpha = 0.3, color = "steelblue") +
  facet_wrap(~ Metric, ncol = 2) +
  geom_hline(data = tibble(
      Metric = paste0("max_", 1:4),
      yintercept = p0.max
    ),
    aes(yintercept = yintercept), linetype = "dashed", color = "red") +
  labs(title = "Convergence of Relative Frequencies", y = "Relative Frequency", x = "Sample Size") +
  theme_minimal()


# 建立理論值資料表
theory_df <- tibble(
  Metric = c("max_mean", "max_var", "max_sd"),
  Theory = c(mean_theory, var_theory, sd_theory)
)

# 繪圖：合併為一張
sim_long %>%
  filter(Metric %in% c("max_mean", "max_var", "max_sd")) %>%
  ggplot(aes(x = SampleSize, y = Value, group = Rep)) +
  geom_line(alpha = 0.3, color = "steelblue") +
  facet_wrap(~ Metric, scales = "free_y", ncol = 1,
             labeller = as_labeller(c(
               max_mean = "Sample Mean",
               max_var = "Sample Variance",
               max_sd = "Sample SD"
             ))) +
  geom_hline(data = theory_df, aes(yintercept = Theory), 
             color = "red", linetype = "dashed") +
  labs(title = "Convergence of Sample Statistics",
       y = "Value", x = "Sample Size") +
  theme_minimal()
```



### **3.1 觀察樣本數變化時的機率估計統計量**

- For a continuous statistic \( h(Y_1, \cdots, Y_n) \), histograms and boxplots serve as useful tools for exploring its distributional properties and potential patterns.

以下是在$n = 50, 100, 500, 1000$ 時，生成上述4個機率估計、樣本平均、樣本變異數、樣本標準差，共 $n\_rep = 1000$ 筆。每一筆或每一列的值，皆來自相同模擬數據所產生。


```{r, echo=FALSE}
p0.max <- (2*(1:4)-1)/16 # 理論機率：Max 的分布（公正骰）

f <- function(n,dicewt){
  # n <- 20
  # dicewt <- c(1,1,1,1)  # 骰子權重
  dicep <- prop.table(dicewt)   # 骰子機率
  p0.ss <- 1/rep(length(dicewt)^2,length(dicewt)^2)  # Sample Space H0:公正骰子
  
  # 生成數據並轉換為 data.frame
  ds <- as.data.frame(matrix(sample(1:4, 2*n, replace=TRUE, prob=dicep), ncol=2))
  colnames(ds) <- c("Var1", "Var2") 
  ds$Max <- pmax(ds$Var1, ds$Var2)
  
  
  # 計算完整的 Sample Space 相對頻率
  fq.ss <- as.vector(table(
    factor(ds$Var1, levels = 1:4),
    factor(ds$Var2, levels = 1:4)
  ))
  chistat.ss <- sum((fq.ss - n * p0.ss)^2 / (n * p0.ss))
  # chisq.test(fq.ss, p = p0.ss) # 卡方適合度檢定
  
  # 計算最大值X的相對頻率
  fq.max <- table(factor(ds$Max,levels=1:4))
  fq.max.rel <- as.numeric(prop.table(fq.max))
  chistat.max <- sum((fq.max-n*p0.max)^2/(n*p0.max))
  # chisq.test(fq.max, p = p0.max) # 卡方適合度檢定

  return(c(
    n = n,
    max_1 = fq.max.rel[1],
    max_2 = fq.max.rel[2],
    max_3 = fq.max.rel[3],
    max_4 = fq.max.rel[4],
    max_mean = mean(ds$Max),
    max_var  = var(ds$Max),
    max_sd   = sd(ds$Max)
    # chistat_ss = chistat.ss,
    # chistat_max = chistat.max
  ))
}

set.seed(1234)
n_rep <- 1000
dicewt <- c(1,1,1,1)
df_stat <- data.frame(rbind(
  t(replicate(n_rep,f(50,dicewt))),
  t(replicate(n_rep,f(100,dicewt))),
  t(replicate(n_rep,f(500,dicewt))),
  t(replicate(n_rep,f(1000,dicewt)))))

# df_stat <- bind_rows(
#   bind_rows(replicate(n_rep, f(20, dicewt), simplify = FALSE)),
#   bind_rows(replicate(n_rep, f(100, dicewt), simplify = FALSE)),
#   bind_rows(replicate(n_rep, f(500, dicewt), simplify = FALSE)),
#   bind_rows(replicate(n_rep, f(500, dicewt), simplify = FALSE))
# )
```


```{r echo=FALSE}
str(df_stat)
```

```{r}
p0.max <- (2 * (1:4) - 1) / 16
names(p0.max) <- paste0("max_", 1:4)  # 對應欄位名稱

# 繪圖
df_stat %>%
  pivot_longer(cols = starts_with("max_"), names_to = "MaxCategory", values_to = "Frequency") %>%
  filter(MaxCategory %in% c("max_1", "max_2", "max_3", "max_4")) %>%
  ggplot(aes(x = factor(n), y = Frequency, fill = MaxCategory, color = MaxCategory)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.05, size = 0.7) +
  geom_hline(data = data.frame(MaxCategory = names(p0.max), y = p0.max),
             aes(yintercept = y, color = MaxCategory), linetype = "dashed", linewidth = 0.6) +
  labs(title = "Boxplot of Relative Frequencies with Raw Points and Theoretical Lines",
       x = "Sample Size (n)", y = "Relative Frequency") +
  theme_minimal()
```

- **Weak Law of Large Numbers (WWLN)**
  Let \( Y_1, Y_2, \ldots \) be i.i.d. random variables with \( E(Y) < \infty \).  
Then for any \( \varepsilon > 0 \),
\[
\lim_{n \to \infty} P\left( \left| \bar{Y}_n - E(Y) \right| \geq \varepsilon \right) = 0,
\]
denoted by 
\[
  \bar{Y}_n \xrightarrow{p} E(Y).
\]

- **Properties of the Sample Mean**
  \( E(\bar{Y}_n) = E(Y) \) (unbiased estimator) and \( \text{Var}(\bar{Y}_n) = \frac{1}{n}\text{Var}(Y) \).
  
  - \(\hat{p}_k = \frac{1}{n} \sum_{i=1}^{n} I_{\{X_i = k\}}\xrightarrow{p} P(X = k).\) (for \(k = 1, 2, 3, 4\))
  
  - \(\text{Var}(\hat{p}_k) = \frac{1}{n} p_k (1 - p_k)\).
  

```{r}
library(tidyverse)

# 長格式
df_long <- df_stat %>%
  pivot_longer(cols = starts_with("max_"), names_to = "MaxCategory", values_to = "Frequency") %>%
  filter(MaxCategory %in% paste0("max_", 1:4))

# 計算每個組別的 mu, sigma
mu_sd <- df_long %>%
  group_by(n, MaxCategory) %>%
  summarise(mu = mean(Frequency), sigma = sd(Frequency), .groups = "drop")

# 建立正態密度資料（用 expand.grid + dnorm）
density_df <- mu_sd %>%
  mutate(xmin = mu - 4 * sigma,
         xmax = mu + 4 * sigma) %>%
  rowwise() %>%
  mutate(x = list(seq(xmin, xmax, length.out = 200)),
         y = list(dnorm(x, mean = mu, sd = sigma))) %>%
  unnest(c(x, y))

# 理論線
p0.df <- tibble(
  MaxCategory = paste0("max_", 1:4),
  Theory = (2 * (1:4) - 1) / 16
)

# 繪圖
ggplot(df_long, aes(x = Frequency)) +
  geom_histogram(aes(y = after_stat(density)), bins = 25,
                 fill = "lightblue", color = "black", # ← 移除邊界線，沒有間距,
                 alpha = 0.7, position = "identity") +
  geom_line(data = density_df, aes(x = x, y = y, color = MaxCategory), linewidth = 0.5, alpha=0.5, inherit.aes = FALSE) +
  geom_vline(data = p0.df, aes(xintercept = Theory, color = MaxCategory),
             linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
  facet_grid(n ~ MaxCategory, scales = "free_x", labeller = label_both) +
  labs(title = "Histogram with Normal Approximation by Sample Size",
       x = "Relative Frequency", y = "Density") +
  theme_minimal()
```



- **Central Limit Theorem (CLT)**:  
  Let \( Y_1, Y_2, \cdots \) be i.i.d. random variables with \( E(Y) < \infty \) and \( \text{Var}(Y) < \infty \). Then, for all \( z \in \mathbb{R} \),
  \[
  \lim_{n \rightarrow \infty} P\left( \frac{\sqrt{n}(\bar{Y}_n - E(Y))}{\sqrt{\text{Var}(Y)}} \leq z \right) = \Phi(z),
  \]
  which is is often denoted as:
  \[
  \frac{\sqrt{n}(\bar{Y}_n - E(Y))}{\sqrt{\text{Var}(Y)}} \xrightarrow{d} N(0, 1).
  \]

- **Example:** By the CLT, we have:
  \[
  \frac{\sqrt{n}(\hat{p}_k - p_k)}{\sqrt{p_k(1 - p_k)}} \xrightarrow{d} N\left( 0, 1\right).  \tag{1}
  \]  
  
  - That is, for large \( n \),
  \[
  \hat{p}_k \approx N\left( p_k, \, \frac{p_k(1 - p_k)}{n} \right)
  \]

  - \(n \hat{p}_k = \sum_{i=1}^{n} I_{\{X_i = k\}} \sim \text{Binomial}(n, p_k).\)
  
  - Moreover, \(\widehat{\text{Var}}(I_{\{X_i = k\}}) = \hat{p}_k (1 - \hat{p}_k) \xrightarrow{p} p_k (1 - p_k), \) by [Convergence of random variables](https://en.wikipedia.org/wiki/Convergence_of_random_variables). Then using [Slutsky's theorem](https://en.wikipedia.org/wiki/Slutsky%27s_theorem), this together with
  \[
  \sqrt{n}(\hat{p}_k  - p_k) \xrightarrow{d} N(0, p_k(1 - p_k))
  \]
  yields
  \[
  \frac{\sqrt{n}( \hat{p}_k - p_k )}{ \sqrt{ \hat{p}_k (1 - \hat{p}_k) } } \xrightarrow{d} N(0, 1).  \tag{2}
  \]
  
  - The sample variance based on \( I_{\{X_i = k\}}, \ i = 1, \ldots, n \), is an unbiased estimator of the true variance \(\text{Var}(I_{\{X_i = k\}}) = p_k(1 - p_k). \) In contrast, 
  \(\widehat{\text{Var}}(I_{\{X_i = k\}}) = \hat{p}_k (1 - \hat{p}_k) \) is a biased estimator, but it is **consistent**. Similarly, following the same reasoning as in Equation (2), we can derive an alternative version of the distributional convergence by substituting the population variance with the sample variance.
  
  
```{r}
library(tidyverse)

for (k in 1:4) {
  var_name <- paste0("max_", k)
  z_theory <- paste0("z_theory_", k)
  z_plugin <- paste0("z_plugin_", k)
  
  # 理論值標準化
  df_stat[[z_theory]] <- sqrt(df_stat$n) * (df_stat[[var_name]] - p0.max[[var_name]]) /
                         sqrt(p0.max[[var_name]] * (1 - p0.max[[var_name]]))
  
  # plug-in 標準化
  df_stat[[z_plugin]] <- sqrt(df_stat$n) * (df_stat[[var_name]] - p0.max[[var_name]]) /
                         sqrt(df_stat[[var_name]] * (1 - df_stat[[var_name]]))
}


library(tidyverse)

# 整成長格式
df_z_long <- df_stat %>%
  pivot_longer(
    cols = starts_with("z_"),
    names_to = c("method", "max_k"),
    names_pattern = "z_(theory|plugin)_(\\d)",
    values_to = "z_value"
  ) %>%
  mutate(
    n = as.factor(n),
    method = if_else(method == "theory", "Z: Theoretical", "Z: Plug-in"),
    max_k = factor(paste0("max_", max_k), levels = paste0("max_", 1:4))
  )

plot_each_max_k <- function(max_k_label) {
  df_z_long %>%
    filter(max_k == max_k_label, is.finite(z_value)) %>%  # 移除 -Inf, Inf, NaN
    ggplot(aes(x = z_value)) +
    geom_histogram(
      aes(y = after_stat(density)),
      bins = 10, fill = "skyblue", color = "white", alpha = 0.7
    ) +
    stat_function(
      fun = dnorm, args = list(mean = 0, sd = 1),
      color = "red", linetype = "dashed", linewidth = 0.8
    ) +
    facet_grid(rows = vars(n), cols = vars(method)) +
    coord_cartesian(xlim = c(-3, 3)) +
    labs(
      title = paste("Standardized Z for", max_k_label),
      x = "Z-value", y = "Density"
    ) +
    theme_minimal() +
    theme(strip.text = element_text(size = 10))
}

```

```{r}
plot_each_max_k("max_1")
# plot_each_max_k("max_2")
# plot_each_max_k("max_3")
# plot_each_max_k("max_4")
```


```{r}
library(dplyr)
library(tidyr)

# Shapiro-Wilk 檢定
shapiro_results <- df_z_long %>%
  filter(is.finite(z_value)) %>%  # 移除 -Inf, Inf, NA
  group_by(n, max_k, method) %>%
  summarise(
    p_value = tryCatch(shapiro.test(z_value)$p.value, error = function(e) NA),
    W = tryCatch(shapiro.test(z_value)$statistic, error = function(e) NA),
    .groups = "drop"
  )

# 分別建立 max_1 ~ max_4 四張表，顯示小數點 4 位並加 * 標記顯著
for (k in 1:4) {
  max_label <- paste0("max_", k)
  
  cat("\n### Shapiro-Wilk p-values for", max_label, "\n\n")
  
  table_k <- shapiro_results %>%
    filter(max_k == max_label) %>%
    select(n, method, p_value) %>%
    pivot_wider(names_from = method, values_from = p_value) %>%
    mutate(across(
      where(is.numeric),
      ~ ifelse(. < 0.05, paste0(format(round(., 4), nsmall = 4), "*"),
                      format(round(., 4), nsmall = 4))
    ))
  
  print(table_k)
}

```


### **3.2 聯合分佈、相關係數**

- Marginal Distribution

- Order Statistics

```{r message=FALSE, warning=FALSE, error=FALSE}
library(GGally)
library(ggplot2)
library(dplyr)

# 整理資料
df_plot <- df_stat %>%
  select(n, max_1, max_2, max_3, max_4) %>%
  mutate(n = factor(n))

# 繪圖（2x2、無 legend、n = xxx 格式）
ggparcoord(
  data = df_plot,
  columns = 2:5,
  groupColumn = 1,
  scale = "globalminmax",
  alphaLines = 0.02,
  showPoints = FALSE
) +
  facet_wrap(
    ~ n,
    nrow = 2,
    labeller = labeller(n = function(x) paste0("n = ", x))
  ) +
  labs(
    title = "Parallel Coordinates of max_1 to max_4 (by n)",
    x = "Max Category", y = "Relative Frequency"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```


```{r message=FALSE, warning=FALSE, error=FALSE}
library(PerformanceAnalytics)
library(dplyr)

# 指定樣本大小
selected_n <- 500

# 資料過濾
chart_data <- df_stat %>%
  filter(n == selected_n) %>%
  select(max_1, max_2, max_3, max_4)

# 畫散佈矩陣圖（包含直方圖與相關係數）
chart.Correlation(chart_data,
                  histogram = TRUE,
                  pch = 16,
                  col = adjustcolor("gray30", alpha.f = 0.5),
                  main = paste("Scatterplot Matrix of max_k (n =", selected_n, ")"))

# 顯示協方差矩陣
cat("The covariance matrix:\n")
print(round(cov(chart_data), 6))


```


```{r}
library(ggplot2)
library(dplyr)

# 設定參數
selected_n <- 100
xvar <- "max_3"
yvar <- "max_4"

# 整理資料
plot_data <- df_stat %>%
  filter(n == selected_n) %>%
  select(all_of(c(xvar, yvar)))

# 根據資料自動取範圍
range_x <- range(plot_data[[xvar]], na.rm = TRUE)
range_y <- range(plot_data[[yvar]], na.rm = TRUE)

# 繪圖
ggplot(plot_data, aes(x = .data[[xvar]], y = .data[[yvar]])) +
  geom_point(alpha = 0.1, color = "gray50") +
  geom_density_2d(color = "blue") +
  geom_density_2d_filled(alpha = 0.4) +
  coord_fixed(xlim = range_x, ylim = range_y) +  # ✅ 自動依據資料決定範圍
  theme_minimal() +
  labs(
    title = paste("2D Density and Contour Plot:", xvar, "vs", yvar),
    x = xvar,
    y = yvar
  )

```
```{r message=FALSE, warning=FALSE, error=FALSE}
library(plotly)
library(dplyr)

# 指定 n
selected_n <- 500

# 選出 max_2 ~ max_4 的資料
chart_data <- df_stat %>%
  filter(n == selected_n) %>%
  select(max_2, max_3, max_4)

# 畫 3D 散佈圖
plot_ly(chart_data, 
        x = ~max_2, 
        y = ~max_3, 
        z = ~max_4,
        type = "scatter3d", 
        mode = "markers",
        marker = list(size = 2, color = 'steelblue', opacity = 0.6)) %>%
  layout(title = paste("3D Scatter Plot of max_2, max_3, max_4 (n =", selected_n, ")"),
         scene = list(
           xaxis = list(title = "max_2"),
           yaxis = list(title = "max_3"),
           zaxis = list(title = "max_4")
         ))

```



### **3.3 樣本平均**

- **Sample Mean**: \( \bar{X}_n \approx N\left( \frac{25}{8}, \, \frac{55}{64n} \right) \)


```{r}
df_stat %>%
  ggplot(aes(x = max_mean)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, fill = "steelblue", color = "white", alpha = 0.7) +
  facet_wrap(~ n, ncol = 2) +
  labs(title = "Histogram of Sample Means by n",
       x = "max_mean", y = "Density") +
  theme_minimal()

shapiro_max_mean <- df_stat %>%
  group_by(n) %>%
  summarise(
    p_value = tryCatch(shapiro.test(max_mean)$p.value, error = function(e) NA),
    .groups = "drop"
  ) %>%
  mutate(
    p_value = ifelse(p_value < 0.05,
                     paste0(format(round(p_value, 4), nsmall = 4), "*"),
                     format(round(p_value, 4), nsmall = 4))
  )

shapiro_max_mean
```

- **Weak Law of Large Numbers (WWLN)**:  


```{r}
library(ggplot2)
library(dplyr)

df_stat %>%
  mutate(n = factor(n)) %>%
  ggplot(aes(x = n, y = max_mean)) +
  geom_boxplot(fill = "skyblue", alpha = 0.7, outlier.shape = NA) +  # 不顯示離群值點
  geom_jitter(width = 0.2, alpha = 0.4, color = "gray30", size = 1) +  # 加上樣本點
  geom_hline(yintercept = 25 / 8, linetype = "dashed", color = "red", linewidth = 0.8) +  # 真實期望值線
  labs(
    title = "Boxplot of Sample Mean by Sample Size with Sample Points",
    x = "Sample Size (n)",
    y = "Sample Mean (max_mean)"
  ) +
  theme_minimal()

```

- **Central Limit Theorem (CLT)** \(E(X)=\frac{25}{8}\) and \(Var(X)=\frac{55}{64}\)
  
  - Using the true variance: \( \frac{\sqrt{n}(\bar{X}_n - E(X))}{\sqrt{\text{Var}(X)}} \xrightarrow{d} N(0, 1) \)

  - Using the sample variance: \( \frac{\sqrt{n}(\bar{X}_n - E(X))}{\sqrt{S^2_n}} \xrightarrow{d} N(0, 1) \)

```{r}
library(tidyverse)

# 理論期望與變異數
mean_theory <- 25 / 8       # 理論平均 = 3.125
mean_theory <- 3.2       # 高估的
var_theory  <- 55 / 64      # 理論變異數 ≈ 0.859375

# 在 df_summary 中新增兩種標準化 Z 值
df_summary <- df_stat %>%
  select(n, max_mean, max_var, max_sd) %>%
  mutate(
    z_truevar = sqrt(n) * (max_mean - mean_theory) / sqrt(var_theory),  # 使用理論變異數
    z_sample  = sqrt(n) * (max_mean - mean_theory) / sqrt(max_var)      # 使用樣本變異數
  )

df_summary %>%
  pivot_longer(
    cols = c(z_truevar, z_sample),
    names_to = "method",
    values_to = "value"
  ) %>%
  mutate(n = factor(n)) %>%
  ggplot(aes(x = value)) +
  geom_histogram(aes(y = after_stat(density), fill = method),
                 bins = 30, alpha = 0.6, position = "identity", color = "white") +
  stat_function(fun = dnorm, args = list(mean = 0, sd = 1),
                color = "red", linetype = "dashed", linewidth = 0.8) +
  facet_grid(rows = vars(n), cols = vars(method)) +
  labs(
    title = "Histogram of Standardized Z-scores (減到不合理的期望值)",
    x = "Z-value", y = "Density", fill = "Method"
  ) +
  theme_minimal()


```

### **3.3 樣本變異數**


- **Sample Variance**:  
  \[
  S_n^2 = \frac{1}{n - 1} \sum_{i = 1}^{n} (Y_i - \bar{Y}_n)^2
  \]
  - \( E(S_n^2) = \sigma^2 \)：an **unbiased estimator** of the population variance  
  
  - \(\text{Var}(S_n^2) = \frac{1}{n} (\kappa - 1 + \frac{2}{n-1}) \sigma^4 \), where \( \kappa \) is the **kurtosis** of the population distribution.
  
    - [wikipedia](https://en.wikipedia.org/wiki/Variance#Sample_variance), [math.stackexchange]( https://math.stackexchange.com/questions/72975/variance-of-sample-variance)
    
    - \(\text{Var}(S_n^2) \approx \frac{1}{n} (\kappa - 1) \sigma^4 \).
  
  - \( S_n^2 \) is a **consistent estimator** of the population variance \( \sigma^2 \); that is, \( S_n^2 \xrightarrow{p} \sigma^2 \) as \( n \to \infty \).
  
  - \( \sqrt{n}(S_n^2 - \sigma^2) \xrightarrow{d} N\left( 0, (\kappa - 1) \sigma^4 \right) \) (試著證看看，亦可用模擬驗證)
  
    - This together with \( \sqrt{(\kappa - 1)} S_n^2 \xrightarrow{p} (\kappa - 1) \sigma^2 \) yields 
    \[
    \frac{\sqrt{n}(S_n^2 - \sigma^2)}{\sqrt{\kappa - 1} S_n^2} \xrightarrow{d} N\left( 0, 1 \right) .
    \]
    
  - For the **normal distribution**, we have:

    - Kurtosis \( \kappa = 3 \), \(\text{Var}(S_n^2) = \frac{2 \sigma^4}{n - 1}\) and \(\frac{(n - 1) S_n^2}{\sigma^2} \sim \chi^2_{n - 1} \).

- A **biased sample variance**:
  \[
  {S'}_n^2 = \frac{1}{n} \sum_{i = 1}^{n} (Y_i - \bar{Y}_n)^2
  \]

  - \(E({S'}_n^2) = \frac{n - 1}{n} \sigma^2 < \sigma^2 \)

  - This estimator is **biased**, but **consistent** as \( n \to \infty \).

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

# 指定 n 為變數
n_val <- 50

# 整理資料
df_hist_n <- df_stat %>%
  filter(n == n_val) %>%
  mutate(adjusted_var = (n - 1) / n * max_var) %>%
  select(max_var, adjusted_var) %>%
  pivot_longer(cols = everything(), names_to = "type", values_to = "value") %>%
  mutate(type = recode(type,
                       max_var = "Sample Variance",
                       adjusted_var = "MLE Variance"))

# 加入 Shapiro-Wilk 正態性檢定的 p-value
shapiro_labels <- df_hist_n %>%
  group_by(type) %>%
  summarise(
    p_value = tryCatch(shapiro.test(value)$p.value, error = function(e) NA_real_),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0("Shapiro p = ", formatC(p_value, digits = 3, format = "f"))
  )

# 合併標籤回原資料
df_hist_n <- df_hist_n %>%
  left_join(shapiro_labels, by = "type")

# 畫圖：加上常態密度 + p-value 標籤
ggplot(df_hist_n, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, fill = "skyblue", color = "white", alpha = 0.7) +
  stat_function(
    fun = dnorm,
    args = list(mean = mean(df_hist_n$value), sd = sd(df_hist_n$value)),
    color = "blue", linetype = "dotted", linewidth = 1
  ) +
  geom_vline(xintercept = 55 / 64, color = "red", linetype = "dashed", linewidth = 0.8) +
  facet_wrap(~ type + label, ncol = 1) +
  labs(
    x = "Value", y = "Density",
    title = paste("Histogram of Variance Estimates (n =", n_val, ")")
  ) +
  theme_minimal()

```

- **Sample Standard Deviation**:  \( S_n = \sqrt{S_n^2} \)

  - \( E(S_n) < \sigma \)：a **biased** estimator of \( \sigma \)

  - For the normal distribution, an **unbiased estimator** of \( \sigma \) is given by:
    \[
    \hat{\sigma} = \frac{S_n}{c_4(n)}
    \quad \text{where } c_4(n) = \sqrt{\frac{2}{n-1}} \cdot \frac{\Gamma(n/2)}{\Gamma((n-1)/2)}
    \]

  - See [Wikipedia: Unbiased estimation of standard deviation](https://en.wikipedia.org/wiki/Unbiased_estimation_of_standard_deviation) for details.


```{r}
library(ggplot2)
library(dplyr)

# 真實的標準差
sd_theory <- sqrt(55 / 64)

# 整理資料
df_sd_plot <- df_stat %>%
  select(n, max_sd) %>%
  mutate(n = factor(n))  # 確保 n 為類別，用於 x 軸

# 畫圖：boxplot + 點圖 + 理論標準差
ggplot(df_sd_plot, aes(x = n, y = max_sd)) +
  geom_boxplot(fill = "lightblue", outlier.shape = NA, width = 0.6) +
  geom_jitter(width = 0.2, alpha = 0.1, color = "darkblue", size = 1) +
  geom_hline(yintercept = sd_theory, color = "red", linetype = "dashed", linewidth = 0.8) +
  labs(
    title = expression("Sample Standard Deviation " ~ S[n] ~ " by Sample Size"),
    x = "Sample Size (n)", y = expression(S[n])
  ) +
  theme_minimal()

c4 <- function(n) {
  sqrt(2 / (n - 1)) * gamma(n / 2) / gamma((n - 1) / 2)
}

# 定義 c4 函數
c4 <- function(n) {
  sqrt(2 / (n - 1)) * gamma(n / 2) / gamma((n - 1) / 2)
}

# 建立 n 與 c4 值的向量
n_vals <- 20:1000
c4_vals <- c4(n_vals)

# 畫圖：使用 base R 的 plot
plot(n_vals, c4_vals,
     type = "l",       # 折線圖
     col = "blue",
     ylim = c(0.987, 1.0001),  # y 軸範圍
     lwd = 2,
     xlab = "n",
     ylab = expression(c[4](n)),
     main = expression(paste("Bias Correction Factor ", c[4](n), " vs n")))

# 加上水平線 y = 1
abline(h = 1, col = "red", lty = 2)

```


### **3.4 樣本平均與樣本變異數不獨立**

```{r}
library(ggplot2)
library(dplyr)

# 理論期望
mean_theory <- 25 / 8     # = 3.125
var_theory  <- 55 / 64    # = 0.859375

# 統計摘要文字
summary_labels <- df_stat %>%
  group_by(n) %>%
  summarise(
    var_mean = var(max_mean),
    var_var  = var(max_var),
    corr     = cor(max_mean, max_var),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0(
      "Var(mean) = ", round(var_mean, 4), "\n",
      "Var(var)  = ", round(var_var, 4), "\n",
      "Corr      = ", round(corr, 4)
    ),
    n = factor(n)
  )

# 合併資料
df_with_label <- df_stat %>%
  mutate(n = factor(n)) %>%
  left_join(summary_labels, by = "n")

# 畫圖：無等高線版本
ggplot(df_with_label, aes(x = max_mean, y = max_var)) +
  geom_point(alpha = 0.3, color = "steelblue", size = 1) +
  geom_vline(xintercept = mean_theory, color = "red", linetype = "dashed") +
  geom_hline(yintercept = var_theory, color = "red", linetype = "dashed") +
  geom_text(data = summary_labels,
            aes(x = -Inf, y = Inf, label = label),
            hjust = -0.1, vjust = 1.1,
            inherit.aes = FALSE,
            size = 3.5, color = "black") +
  facet_wrap(~ n) +
  labs(
    title = "Scatter Plot of Sample Mean vs. Variance by Sample Size",
    x = expression(bar(X)[n]),
    y = expression(S[n]^2)
  ) +
  theme_minimal()

```


- https://chatgpt.com/canvas/shared/67ea3d706d4081919bf5129e7e6aa708

**問題思考**: 若骰子不公平，那麼:

  - \( P(X = k) \) 會是什麼樣的分佈呢？期望值、變異數又會是什麼樣的數值呢？
  
  - 除了手算外，你還可以怎麼算？
